---
title: "HPC-VS Connectivity and Past-Year 5+ Drinking Frequency"
author: "Blake Elliott"
date: "2026-05-31"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
params:
  data_path: !r Sys.getenv("HPC_DATA_PATH", unset = "averaged_data.csv")
  n_perm: 10000
  bootstrap_reps: 500
  run_permutations: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.pos = "htbp",
  out.width = "100%",
  dpi = 150,
  fig.path = file.path(tempdir(), "hpc-report-figures-")
)

DATA_PATH <- params$data_path
if (!file.exists(DATA_PATH)) {
  stop(
    "Data file not found. Put averaged_data.csv beside this Rmd file or set ",
    "the HPC_DATA_PATH environment variable before knitting."
  )
}
```

# HPC-VS Binge Drinking Analysis

This R Markdown file runs the analyses and creates the figures reported in the
manuscript and supplement:

1. Primary proportional-odds ordinal logistic models.
2. Permutation likelihood-ratio checks with 10,000 permutations.
3. DFBETAS influence sensitivity checks.
4. Sex-moderation models and conditional slopes.
5. Age sensitivity analyses.
6. Three-level alcohol-diagnosis follow-ups: no lifetime DSM-IV alcohol diagnosis,
   lifetime DSM-IV alcohol abuse, and lifetime DSM-IV alcohol dependence.
7. Diagnosis-group ANCOVAs for standardized left and right HPC-VS connectivity.
8. Manuscript Figures 2-5.


- `set.seed(23)`
- `N_PERM <- 10000`
- DFBETAS screens include max `|DFBETAS| > .20` and `2/sqrt(N)`

```{r analysis-setup, include=FALSE}
options(contrasts = c("contr.sum", "contr.poly"))
set.seed(23)

suppressPackageStartupMessages({
  library(tidyverse)
  library(MASS)
  library(rio)
  library(emmeans)
})

PERM_SEED <- 23
N_PERM <- params$n_perm
BOOT_REPS <- params$bootstrap_reps
RMS_DFBETAS_CUTOFF <- 0.20

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), "NA", formatC(x, format = "f", digits = digits))
}

fmt_p <- function(p) {
  ifelse(
    is.na(p),
    "NA",
    ifelse(p < 0.001, "< .001", sub("^0", "", formatC(p, format = "f", digits = 3)))
  )
}

fit_polr_safe <- function(formula, data) {
  suppressWarnings(
    MASS::polr(formula, data = data, Hess = TRUE, method = "logistic", model = TRUE)
  )
}

coef_table <- function(model) {
  ct <- coef(summary(model))
  p <- 2 * pnorm(abs(ct[, "t value"]), lower.tail = FALSE)

  tibble(
    term = rownames(ct),
    b = ct[, "Value"],
    SE = ct[, "Std. Error"],
    z = ct[, "t value"],
    p = p,
    OR = exp(ct[, "Value"])
  )
}

lr_stat <- function(reduced_model, full_model) {
  2 * (as.numeric(logLik(full_model)) - as.numeric(logLik(reduced_model)))
}

lr_test <- function(reduced_model, full_model, label) {
  lr <- lr_stat(reduced_model, full_model)
  df <- attr(logLik(full_model), "df") - attr(logLik(reduced_model), "df")
  tibble(
    test = label,
    df = df,
    LR = lr,
    p = pchisq(lr, df = df, lower.tail = FALSE)
  )
}

lm_block_test <- function(reduced_model, full_model, label) {
  a <- anova(reduced_model, full_model)
  tibble(
    test = label,
    df_num = a$Df[2],
    df_den = a$Res.Df[2],
    F = a$F[2],
    p = a$`Pr(>F)`[2]
  )
}

lm_coef_table <- function(model) {
  ct <- coef(summary(model))
  tibble(
    term = rownames(ct),
    b = ct[, "Estimate"],
    SE = ct[, "Std. Error"],
    t = ct[, "t value"],
    p = ct[, "Pr(>|t|)"]
  )
}

fmt_count_pct <- function(n, denom) {
  paste0(n, " (", fmt_num(100 * n / denom, 1), "%)")
}

perm_lr_test <- function(reduced_formula, full_formula, data, label, n_perm = N_PERM, seed = PERM_SEED) {
  set.seed(seed)

  reduced_obs <- fit_polr_safe(reduced_formula, data)
  full_obs <- fit_polr_safe(full_formula, data)
  observed_lr <- lr_stat(reduced_obs, full_obs)

  perm_stats <- replicate(n_perm, {
    dat_p <- data
    dat_p$y <- sample(dat_p$y, replace = FALSE)

    val <- try({
      red_p <- fit_polr_safe(reduced_formula, dat_p)
      full_p <- fit_polr_safe(full_formula, dat_p)
      lr_stat(red_p, full_p)
    }, silent = TRUE)

    if (inherits(val, "try-error") || !is.finite(val)) NA_real_ else val
  })

  perm_stats <- perm_stats[is.finite(perm_stats)]
  p_perm <- (sum(perm_stats >= observed_lr) + 1) / (length(perm_stats) + 1)

  tibble(
    test = label,
    LR = observed_lr,
    p_perm = p_perm,
    n_success = length(perm_stats)
  )
}

dfbetas_screen <- function(formula, data, label) {
  full_model <- fit_polr_safe(formula, data)
  full_beta <- coef(full_model)
  se_lookup <- coef_table(full_model) %>%
    filter(.data$term %in% names(full_beta)) %>%
    dplyr::select(term, SE)

  purrr::map_dfr(seq_len(nrow(data)), function(i) {
    deleted_model <- fit_polr_safe(formula, droplevels(data[-i, , drop = FALSE]))
    deleted_beta <- coef(deleted_model)[names(full_beta)]

    tibble(
      model = label,
      row_id = data$row_id[i],
      Subject = data$Subject[i],
      deleted_position = i,
      term = names(full_beta),
      dfbetas = as.numeric((deleted_beta - full_beta) / se_lookup$SE)
    )
  }) %>%
    mutate(abs_dfbetas = abs(dfbetas))
}

summarize_dfbetas_cases <- function(screen_long, cutoff, cutoff_label) {
  screen_long %>%
    mutate(flagged_term = abs_dfbetas > cutoff) %>%
    group_by(model, row_id, Subject) %>%
    summarize(
      cutoff_label = cutoff_label,
      cutoff = cutoff,
      max_abs_dfbetas = max(abs_dfbetas, na.rm = TRUE),
      flagged_any = any(flagged_term),
      flagged_terms = paste(term[flagged_term], collapse = "; "),
      .groups = "drop"
    ) %>%
    arrange(desc(max_abs_dfbetas))
}

dfbetas_refit <- function(label, data, reduced_formula, full_formula, target_term, screen_long, cutoff, cutoff_label) {
  cases <- summarize_dfbetas_cases(screen_long, cutoff, cutoff_label)
  flagged <- cases %>% filter(flagged_any)
  filtered_data <- data %>% filter(!row_id %in% flagged$row_id) %>% droplevels()

  reduced_full_sample <- fit_polr_safe(reduced_formula, data)
  full_full_sample <- fit_polr_safe(full_formula, data)
  full_coef <- coef_table(full_full_sample) %>% filter(.data$term == !!target_term)
  full_lr <- lr_test(reduced_full_sample, full_full_sample, "Full sample")

  reduced_filtered <- fit_polr_safe(reduced_formula, filtered_data)
  full_filtered <- fit_polr_safe(full_formula, filtered_data)
  filtered_coef <- coef_table(full_filtered) %>% filter(.data$term == !!target_term)
  filtered_lr <- lr_test(reduced_filtered, full_filtered, cutoff_label)

  result <- tibble(
    model = label,
    cutoff_label = cutoff_label,
    cutoff = cutoff,
    removed_n = nrow(flagged),
    retained_n = nrow(filtered_data),
    target_term = target_term,
    b = filtered_coef$b,
    SE = filtered_coef$SE,
    z = filtered_coef$z,
    p = filtered_coef$p,
    OR = filtered_coef$OR,
    LR = filtered_lr$LR,
    LR_df = filtered_lr$df,
    LR_p = filtered_lr$p
  )

  list(
    cases = cases,
    flagged = flagged,
    filtered_data = filtered_data,
    filtered_model = full_filtered,
    result = result,
    full_result = tibble(
      model = label,
      cutoff_label = "Full sample",
      cutoff = NA_real_,
      removed_n = 0L,
      retained_n = nrow(data),
      target_term = target_term,
      b = full_coef$b,
      SE = full_coef$SE,
      z = full_coef$z,
      p = full_coef$p,
      OR = full_coef$OR,
      LR = full_lr$LR,
      LR_df = full_lr$df,
      LR_p = full_lr$p
    )
  )
}

run_dfbetas_checks <- function(label, data, reduced_formula, full_formula, target_term) {
  bkw_cutoff <- 2 / sqrt(nrow(data))
  screen_long <- dfbetas_screen(full_formula, data, label)

  rms <- dfbetas_refit(
    label = label,
    data = data,
    reduced_formula = reduced_formula,
    full_formula = full_formula,
    target_term = target_term,
    screen_long = screen_long,
    cutoff = RMS_DFBETAS_CUTOFF,
    cutoff_label = "max |DFBETAS| > .20"
  )

  bkw <- dfbetas_refit(
    label = label,
    data = data,
    reduced_formula = reduced_formula,
    full_formula = full_formula,
    target_term = target_term,
    screen_long = screen_long,
    cutoff = bkw_cutoff,
    cutoff_label = paste0("max |DFBETAS| > 2/sqrt(N) = ", fmt_num(bkw_cutoff, 4))
  )

  result <- bind_rows(rms$full_result, rms$result, bkw$result)

  list(screen_long = screen_long, rms = rms, bkw = bkw, result = result)
}

simple_slopes <- function(model, var_name, label) {
  trends <- emmeans::emtrends(model, specs = ~ Sex, var = var_name, mode = "latent")
  slopes <- as.data.frame(summary(trends, infer = c(TRUE, TRUE))) %>%
    as_tibble() %>%
    transmute(
      analysis = label,
      contrast = as.character(Sex),
      b = !!sym(paste0(var_name, ".trend")),
      SE = SE,
      z = z.ratio,
      p = p.value,
      lower.CL = asymp.LCL,
      upper.CL = asymp.UCL
    )

  diff <- as.data.frame(summary(pairs(trends), infer = c(TRUE, TRUE))) %>%
    as_tibble() %>%
    transmute(
      analysis = label,
      contrast = as.character(contrast),
      b = estimate,
      SE = SE,
      z = z.ratio,
      p = p.value,
      lower.CL = asymp.LCL,
      upper.CL = asymp.UCL
    )

  bind_rows(slopes, diff)
}

HCP_dat <- rio::import(DATA_PATH)

analysis_dat <- HCP_dat %>%
  mutate(
    Subject = as.character(Subject),
    Sex = factor(trimws(Gender), levels = c("F", "M")),
    binge_freq_rc = case_when(
      SSAGA_Alc_12_Frq_5plus == 5 ~ 1L,
      SSAGA_Alc_12_Frq_5plus == 4 ~ 2L,
      SSAGA_Alc_12_Frq_5plus == 3 ~ 3L,
      SSAGA_Alc_12_Frq_5plus %in% c(1L, 2L) ~ 4L,
      TRUE ~ NA_integer_
    ),
    binge_freq_rc_f = factor(
      binge_freq_rc,
      levels = c(1, 2, 3, 4),
      labels = c("Never", "1-11 d/year", "1-3 d/month", "Weekly or greater"),
      ordered = TRUE
    ),
    alcohol_abuse_dx = case_when(
      SSAGA_Alc_D4_Ab_Dx == 5 ~ TRUE,
      SSAGA_Alc_D4_Ab_Dx == 1 ~ FALSE,
      TRUE ~ NA
    ),
    alcohol_dependence_dx = case_when(
      SSAGA_Alc_D4_Dp_Dx == 5 ~ TRUE,
      SSAGA_Alc_D4_Dp_Dx == 1 ~ FALSE,
      TRUE ~ NA
    ),
    AUD = factor(
      if_else(alcohol_abuse_dx | alcohol_dependence_dx, "AUD", "No AUD"),
      levels = c("No AUD", "AUD")
    ),
    DxGroup = factor(
      case_when(
        alcohol_dependence_dx ~ "Dependence",
        alcohol_abuse_dx ~ "Abuse",
        alcohol_abuse_dx == FALSE & alcohol_dependence_dx == FALSE ~ "No AUD",
        TRUE ~ NA_character_
      ),
      levels = c("No AUD", "Abuse", "Dependence")
    )
  ) %>%
  transmute(
    row_id = row_number(),
    Subject,
    Sex,
    y = binge_freq_rc_f,
    alcohol_abuse_dx,
    alcohol_dependence_dx,
    AUD,
    DxGroup,
    left_hpc_vs = as.numeric(limbic_Left),
    right_hpc_vs = as.numeric(limbic_Right),
    ICV = as.numeric(FS_IntraCranial_Vol),
    Age = as.numeric(Age_in_Yrs)
  ) %>%
  filter(
    is.finite(left_hpc_vs),
    is.finite(right_hpc_vs),
    is.finite(ICV),
    is.finite(Age),
    !is.na(y),
    !is.na(Sex),
    !is.na(AUD),
    !is.na(DxGroup)
  ) %>%
  mutate(
    c_left_hpc_vs = as.numeric(scale(left_hpc_vs)),
    c_right_hpc_vs = as.numeric(scale(right_hpc_vs)),
    c_ICV = as.numeric(scale(ICV)),
    c_Age = as.numeric(scale(Age))
  ) %>%
  droplevels()

contrasts(analysis_dat$AUD) <- matrix(
  c(0, 1),
  ncol = 1,
  dimnames = list(levels(analysis_dat$AUD), "AUD")
)
contrasts(analysis_dat$DxGroup) <- matrix(
  c(0, 1, 0, 0, 0, 1),
  ncol = 2,
  dimnames = list(levels(analysis_dat$DxGroup), c("Abuse", "Dependence"))
)

left_dat <- analysis_dat %>%
  dplyr::select(row_id, Subject, Sex, y, c_left_hpc_vs, c_ICV, c_Age) %>%
  droplevels()

right_dat <- analysis_dat %>%
  dplyr::select(row_id, Subject, Sex, y, c_right_hpc_vs, c_ICV, c_Age) %>%
  droplevels()
```

# Statistical Models and Analyses

```{r permutation-results-setup, include=FALSE}
precomputed_permutation <- tibble::tribble(
  ~test, ~LR, ~p_perm, ~n_success,
  "Permutation LR, left main effect", 4.282, 0.036, 10000L,
  "Permutation LR, right main effect", 0.034, 0.850, 10000L,
  "Permutation LR, left model add sex", 13.831, 1 / 10001, 10000L,
  "Permutation LR, left interaction", 0.000, 0.985, 10000L,
  "Permutation LR, right model add sex", 14.604, 1 / 10001, 10000L,
  "Permutation LR, right interaction", 3.001, 0.082, 10000L
)

get_permutation_result <- function(reduced_formula, full_formula, data, label) {
  if (isTRUE(params$run_permutations)) {
    perm_lr_test(reduced_formula, full_formula, data, label)
  } else {
    precomputed_permutation %>%
      dplyr::filter(.data$test == label)
  }
}
```

## Main-Effect Models

### Left Hemisphere

```{r left-main-model, echo=TRUE}
left_base <- fit_polr_safe(y ~ c_ICV, data = left_dat)
left_main <- fit_polr_safe(y ~ c_ICV + c_left_hpc_vs, data = left_dat)

summary(left_main)
anova(left_base, left_main)

left_main_lr <- lr_test(
  left_base,
  left_main,
  "Left main effect: add HPC-VS connectivity"
)
left_main_lr
```

```{r left-main-permutation, echo=TRUE}
left_main_perm <- get_permutation_result(
  y ~ c_ICV,
  y ~ c_ICV + c_left_hpc_vs,
  left_dat,
  "Permutation LR, left main effect"
)
left_main_perm
```

```{r left-main-influence, echo=TRUE}
left_main_dfb <- run_dfbetas_checks(
  "left_main",
  left_dat,
  y ~ c_ICV,
  y ~ c_ICV + c_left_hpc_vs,
  "c_left_hpc_vs"
)
left_main_dfb$result
```

### Right Hemisphere

```{r right-main-model, echo=TRUE}
right_base <- fit_polr_safe(y ~ c_ICV, data = right_dat)
right_main <- fit_polr_safe(y ~ c_ICV + c_right_hpc_vs, data = right_dat)

summary(right_main)
anova(right_base, right_main)

right_main_lr <- lr_test(
  right_base,
  right_main,
  "Right main effect: add HPC-VS connectivity"
)
right_main_lr
```

```{r right-main-permutation, echo=TRUE}
right_main_perm <- get_permutation_result(
  y ~ c_ICV,
  y ~ c_ICV + c_right_hpc_vs,
  right_dat,
  "Permutation LR, right main effect"
)
right_main_perm
```

```{r right-main-influence, echo=TRUE}
right_main_dfb <- run_dfbetas_checks(
  "right_main",
  right_dat,
  y ~ c_ICV,
  y ~ c_ICV + c_right_hpc_vs,
  "c_right_hpc_vs"
)
right_main_dfb$result
```

## Sex-Moderation Models

### Left Hemisphere

```{r left-sex-moderation, echo=TRUE}
left_sex <- fit_polr_safe(y ~ c_ICV + c_left_hpc_vs + Sex, data = left_dat)
left_int <- fit_polr_safe(y ~ c_ICV + c_left_hpc_vs * Sex, data = left_dat)

summary(left_int)
anova(left_main, left_sex, left_int)

left_sex_lr <- lr_test(left_main, left_sex, "Left model: add sex")
left_int_lr <- lr_test(
  left_sex,
  left_int,
  "Left model: add HPC-VS connectivity x Sex"
)
left_sex_lr
left_int_lr
```

```{r left-sex-permutations, echo=TRUE}
left_sex_perm <- get_permutation_result(
  y ~ c_ICV + c_left_hpc_vs,
  y ~ c_ICV + c_left_hpc_vs + Sex,
  left_dat,
  "Permutation LR, left model add sex"
)
left_int_perm <- get_permutation_result(
  y ~ c_ICV + c_left_hpc_vs + Sex,
  y ~ c_ICV + c_left_hpc_vs * Sex,
  left_dat,
  "Permutation LR, left interaction"
)
left_sex_perm
left_int_perm
```

```{r left-sex-influence-and-slopes, echo=TRUE}
left_int_dfb <- run_dfbetas_checks(
  "left_interaction",
  left_dat,
  y ~ c_ICV + c_left_hpc_vs + Sex,
  y ~ c_ICV + c_left_hpc_vs * Sex,
  "c_left_hpc_vs:Sex1"
)
left_int_dfb$result

emmeans::emtrends(
  left_int,
  specs = ~ Sex,
  var = "c_left_hpc_vs",
  mode = "latent"
)

left_slopes_full <- simple_slopes(left_int, "c_left_hpc_vs", "Full sample")
left_slopes_rms <- simple_slopes(
  left_int_dfb$rms$filtered_model,
  "c_left_hpc_vs",
  "After max |DFBETAS| > .20 removal"
)
left_slopes_bkw <- simple_slopes(
  left_int_dfb$bkw$filtered_model,
  "c_left_hpc_vs",
  "After max |DFBETAS| > 2/sqrt(N) removal"
)
left_slopes_full
left_slopes_rms
left_slopes_bkw
```

### Right Hemisphere

```{r right-sex-moderation, echo=TRUE}
right_sex <- fit_polr_safe(y ~ c_ICV + c_right_hpc_vs + Sex, data = right_dat)
right_int <- fit_polr_safe(y ~ c_ICV + c_right_hpc_vs * Sex, data = right_dat)

summary(right_int)
anova(right_main, right_sex, right_int)

right_sex_lr <- lr_test(right_main, right_sex, "Right model: add sex")
right_int_lr <- lr_test(
  right_sex,
  right_int,
  "Right model: add HPC-VS connectivity x Sex"
)
right_sex_lr
right_int_lr
```

```{r right-sex-permutations, echo=TRUE}
right_sex_perm <- get_permutation_result(
  y ~ c_ICV + c_right_hpc_vs,
  y ~ c_ICV + c_right_hpc_vs + Sex,
  right_dat,
  "Permutation LR, right model add sex"
)
right_int_perm <- get_permutation_result(
  y ~ c_ICV + c_right_hpc_vs + Sex,
  y ~ c_ICV + c_right_hpc_vs * Sex,
  right_dat,
  "Permutation LR, right interaction"
)
right_sex_perm
right_int_perm
```

```{r right-sex-influence-and-slopes, echo=TRUE}
right_int_dfb <- run_dfbetas_checks(
  "right_interaction",
  right_dat,
  y ~ c_ICV + c_right_hpc_vs + Sex,
  y ~ c_ICV + c_right_hpc_vs * Sex,
  "c_right_hpc_vs:Sex1"
)
right_int_dfb$result

emmeans::emtrends(
  right_int,
  specs = ~ Sex,
  var = "c_right_hpc_vs",
  mode = "latent"
)

right_slopes_full <- simple_slopes(right_int, "c_right_hpc_vs", "Full sample")
right_slopes_rms <- simple_slopes(
  right_int_dfb$rms$filtered_model,
  "c_right_hpc_vs",
  "After max |DFBETAS| > .20 removal"
)
right_slopes_bkw <- simple_slopes(
  right_int_dfb$bkw$filtered_model,
  "c_right_hpc_vs",
  "After max |DFBETAS| > 2/sqrt(N) removal"
)
right_slopes_full
right_slopes_rms
right_slopes_bkw
```

## Age Sensitivity Models

```{r age-sensitivity-models, echo=TRUE}
left_age_reduced <- fit_polr_safe(y ~ c_ICV + c_Age, data = left_dat)
left_age_full <- fit_polr_safe(
  y ~ c_ICV + c_Age + c_left_hpc_vs,
  data = left_dat
)
summary(left_age_full)
anova(left_age_reduced, left_age_full)
left_age_lr <- lr_test(
  left_age_reduced,
  left_age_full,
  "Left + age: add HPC-VS connectivity"
)

right_age_reduced <- fit_polr_safe(
  y ~ c_ICV + c_Age + c_right_hpc_vs + Sex,
  data = right_dat
)
right_age_full <- fit_polr_safe(
  y ~ c_ICV + c_Age + c_right_hpc_vs * Sex,
  data = right_dat
)
summary(right_age_full)
anova(right_age_reduced, right_age_full)
right_age_lr <- lr_test(
  right_age_reduced,
  right_age_full,
  "Right + age: add HPC-VS connectivity x Sex"
)

left_age_mod_reduced <- fit_polr_safe(
  y ~ c_ICV + c_Age + c_left_hpc_vs,
  data = left_dat
)
left_age_mod_full <- fit_polr_safe(
  y ~ c_ICV + c_Age * c_left_hpc_vs,
  data = left_dat
)
summary(left_age_mod_full)
anova(left_age_mod_reduced, left_age_mod_full)
left_age_mod_lr <- lr_test(
  left_age_mod_reduced,
  left_age_mod_full,
  "Left: add HPC-VS connectivity x Age"
)

right_age_mod_reduced <- fit_polr_safe(
  y ~ c_ICV + c_Age + c_right_hpc_vs * Sex,
  data = right_dat
)
right_age_mod_full <- fit_polr_safe(
  y ~ c_ICV + c_Age + c_right_hpc_vs * Sex +
    c_right_hpc_vs:c_Age + Sex:c_Age +
    c_right_hpc_vs:Sex:c_Age,
  data = right_dat
)
summary(right_age_mod_full)
anova(right_age_mod_reduced, right_age_mod_full)
right_age_mod_lr <- lr_test(
  right_age_mod_reduced,
  right_age_mod_full,
  "Right: add age moderation terms"
)

left_age_lr
right_age_lr
left_age_mod_lr
right_age_mod_lr
```

## Alcohol Diagnosis Analyses

```{r diagnosis-descriptives, echo=TRUE}
sex_totals <- table(analysis_dat$Sex)
total_n <- nrow(analysis_dat)
female_n <- unname(sex_totals["F"])
male_n <- unname(sex_totals["M"])
female_abuse <- sum(analysis_dat$Sex == "F" & analysis_dat$alcohol_abuse_dx)
male_abuse <- sum(analysis_dat$Sex == "M" & analysis_dat$alcohol_abuse_dx)
total_abuse <- sum(analysis_dat$alcohol_abuse_dx)
female_dependence <- sum(analysis_dat$Sex == "F" & analysis_dat$alcohol_dependence_dx)
male_dependence <- sum(analysis_dat$Sex == "M" & analysis_dat$alcohol_dependence_dx)
total_dependence <- sum(analysis_dat$alcohol_dependence_dx)
female_aud <- sum(analysis_dat$Sex == "F" & analysis_dat$AUD == "AUD")
male_aud <- sum(analysis_dat$Sex == "M" & analysis_dat$AUD == "AUD")
total_aud <- sum(analysis_dat$AUD == "AUD")

diagnosis_counts <- tibble(
  characteristic = c(
    "Sample size",
    "Alcohol abuse",
    "Alcohol dependence",
    "Any alcohol use disorder"
  ),
  female_n = c(female_n, female_abuse, female_dependence, female_aud),
  male_n = c(male_n, male_abuse, male_dependence, male_aud),
  total_n = c(total_n, total_abuse, total_dependence, total_aud)
)

diagnosis_counts
table(analysis_dat$DxGroup)
table(analysis_dat$Sex, analysis_dat$DxGroup)
```

```{r diagnosis-moderation-models, echo=TRUE}
left_dx_mod_reduced <- fit_polr_safe(
  y ~ c_ICV + DxGroup + c_left_hpc_vs,
  data = analysis_dat
)
left_dx_mod_full <- fit_polr_safe(
  y ~ c_ICV + DxGroup * c_left_hpc_vs,
  data = analysis_dat
)
summary(left_dx_mod_full)
anova(left_dx_mod_reduced, left_dx_mod_full)
left_dx_mod_lr <- lr_test(
  left_dx_mod_reduced,
  left_dx_mod_full,
  "Left: add diagnosis group x HPC-VS connectivity"
)

right_dx_mod_reduced <- fit_polr_safe(
  y ~ c_ICV + DxGroup + c_right_hpc_vs * Sex,
  data = analysis_dat
)
right_dx_mod_full <- fit_polr_safe(
  y ~ c_ICV + c_right_hpc_vs * Sex * DxGroup,
  data = analysis_dat
)
summary(right_dx_mod_full)
anova(right_dx_mod_reduced, right_dx_mod_full)
right_dx_mod_lr <- lr_test(
  right_dx_mod_reduced,
  right_dx_mod_full,
  "Right: add diagnosis-group moderation terms"
)

left_dx_mod_lr
right_dx_mod_lr
```

```{r diagnosis-connectivity-models, echo=TRUE}
left_conn_reduced <- lm(c_left_hpc_vs ~ Sex + c_ICV, data = analysis_dat)
left_conn_full <- lm(
  c_left_hpc_vs ~ DxGroup + Sex + c_ICV,
  data = analysis_dat
)
right_conn_reduced <- lm(c_right_hpc_vs ~ Sex + c_ICV, data = analysis_dat)
right_conn_full <- lm(
  c_right_hpc_vs ~ DxGroup + Sex + c_ICV,
  data = analysis_dat
)

summary(left_conn_full)
anova(left_conn_reduced, left_conn_full)
summary(right_conn_full)
anova(right_conn_reduced, right_conn_full)

left_conn_ancova <- lm_block_test(
  left_conn_reduced,
  left_conn_full,
  "Left HPC-VS connectivity: add diagnosis group"
)
right_conn_ancova <- lm_block_test(
  right_conn_reduced,
  right_conn_full,
  "Right HPC-VS connectivity: add diagnosis group"
)

left_conn_emm <- emmeans::emmeans(left_conn_full, ~ DxGroup)
right_conn_emm <- emmeans::emmeans(right_conn_full, ~ DxGroup)

left_conn_emm
right_conn_emm
pairs(left_conn_emm)
pairs(right_conn_emm)

diagnosis_connectivity_emmeans <- bind_rows(
  as_tibble(as.data.frame(summary(left_conn_emm))) %>%
    mutate(hemisphere = "Left", .before = 1),
  as_tibble(as.data.frame(summary(right_conn_emm))) %>%
    mutate(hemisphere = "Right", .before = 1)
)
diagnosis_connectivity_pairwise <- bind_rows(
  as_tibble(as.data.frame(summary(pairs(left_conn_emm), infer = c(TRUE, TRUE)))) %>%
    mutate(hemisphere = "Left", .before = 1),
  as_tibble(as.data.frame(summary(pairs(right_conn_emm), infer = c(TRUE, TRUE)))) %>%
    mutate(hemisphere = "Right", .before = 1)
)

diagnosis_connectivity_emmeans
diagnosis_connectivity_pairwise
```

```{r collect-analysis-results, include=FALSE}
results_primary <- bind_rows(
  left_main_lr,
  right_main_lr,
  left_sex_lr,
  left_int_lr,
  right_sex_lr,
  right_int_lr
)

results_permutation <- bind_rows(
  left_main_perm,
  right_main_perm,
  left_sex_perm,
  left_int_perm,
  right_sex_perm,
  right_int_perm
)

results_dfbetas <- bind_rows(
  left_main_dfb$result,
  right_main_dfb$result,
  left_int_dfb$result,
  right_int_dfb$result
)

results_slopes <- bind_rows(
  left_slopes_full %>% mutate(hemisphere = "Left", .before = 1),
  left_slopes_rms %>% mutate(hemisphere = "Left", .before = 1),
  left_slopes_bkw %>% mutate(hemisphere = "Left", .before = 1),
  right_slopes_full %>% mutate(hemisphere = "Right", .before = 1),
  right_slopes_rms %>% mutate(hemisphere = "Right", .before = 1),
  right_slopes_bkw %>% mutate(hemisphere = "Right", .before = 1)
)

results_age_sensitivity <- bind_rows(
  left_age_lr,
  right_age_lr,
  left_age_mod_lr,
  right_age_mod_lr
)

results_diagnosis_moderation <- bind_rows(
  left_dx_mod_lr,
  right_dx_mod_lr
)

results_diagnosis_ancova <- bind_rows(
  left_conn_ancova,
  right_conn_ancova
)
```

# Final Figure 2

This section displays the sex-stratified drinking-frequency distribution figure
directly in the report.

\setcounter{figure}{1}

```{r final-figure-2, fig.width=10.5, fig.height=6.4, fig.cap="Distribution of past-year 5+ drinking frequency by sex."}
options(contrasts = c("contr.sum", "contr.poly"))

suppressPackageStartupMessages({
  library(tidyverse)
  library(rio)
  library(ggplot2)
  library(scales)
})

HCP_dat <- rio::import(DATA_PATH)

analysis_dat <- HCP_dat %>%
  mutate(
    Subject = as.character(Subject),
    Sex = factor(trimws(Gender), levels = c("F", "M")),
    binge_freq_rc = case_when(
      SSAGA_Alc_12_Frq_5plus == 5 ~ 1L,
      SSAGA_Alc_12_Frq_5plus == 4 ~ 2L,
      SSAGA_Alc_12_Frq_5plus == 3 ~ 3L,
      SSAGA_Alc_12_Frq_5plus %in% c(1L, 2L) ~ 4L,
      TRUE ~ NA_integer_
    ),
    binge_freq_rc_f = factor(
      binge_freq_rc,
      levels = c(1, 2, 3, 4),
      labels = c("Never", "1-11 d/year", "1-3 d/month", "Weekly or greater"),
      ordered = TRUE
    )
  ) %>%
  transmute(
    Subject,
    Sex,
    y = binge_freq_rc_f,
    left_hpc_vs = as.numeric(limbic_Left),
    right_hpc_vs = as.numeric(limbic_Right),
    ICV = as.numeric(FS_IntraCranial_Vol),
    Age = as.numeric(Age_in_Yrs)
  ) %>%
  filter(
    is.finite(left_hpc_vs),
    is.finite(right_hpc_vs),
    is.finite(ICV),
    is.finite(Age),
    !is.na(y),
    !is.na(Sex)
  ) %>%
  droplevels()

plot_dat_dist <- analysis_dat %>%
  count(Sex, y, name = "n") %>%
  group_by(Sex) %>%
  mutate(
    prop = n / sum(n),
    label = if_else(
      prop < .13,
      paste0(n, " (", percent(prop, accuracy = 1), ")"),
      paste0(n, "\n(", percent(prop, accuracy = 1), ")")
    ),
    text_color = if_else(y == "Weekly or greater", "white", "black")
  ) %>%
  ungroup()

fig_dist <- ggplot(plot_dat_dist, aes(x = Sex, y = prop, fill = y)) +
  geom_col(width = 0.68, color = "white", linewidth = 1.0) +
  geom_text(
    aes(label = label, color = text_color),
    position = position_stack(vjust = 0.5),
    size = 6.0,
    lineheight = 0.88,
    fontface = "bold"
  ) +
  scale_color_identity() +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(
    values = c(
      "Never" = "#dbeafe",
      "1-11 d/year" = "#93c5fd",
      "1-3 d/month" = "#3b82f6",
      "Weekly or greater" = "#1e3a8a"
    )
  ) +
  labs(
    x = "Sex",
    y = "Percent within sex",
    fill = "Past-year 5+ drinking frequency",
    title = "Past-year 5+ drinking frequency\nby sex"
  ) +
  theme_bw(base_size = 24) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 26, lineheight = 0.94, margin = margin(b = 12)),
    axis.title = element_text(size = 27),
    axis.text = element_text(size = 25, color = "gray15"),
    legend.title = element_text(size = 23),
    legend.text = element_text(size = 22),
    legend.key.height = unit(0.34, "in"),
    legend.key.width = unit(0.44, "in"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray88", linewidth = 0.7),
    plot.margin = margin(18, 24, 18, 24)
  )

fig_dist
```

```{r figure-2-count-table}
plot_dat_dist %>%
  dplyr::select(Sex, y, n, prop)
```

# Final Figures 3 And 4

This section displays the pooled and sex-stratified HPC-VS ordinal association
figures with bootstrap confidence ribbons.

```{r final-figures-3-4, fig.width=12.5, fig.height=7.1, results='asis', fig.cap=c("Pooled HPC-VS ordinal associations.", "Sex-stratified HPC-VS ordinal associations.")}
options(contrasts = c("contr.sum", "contr.poly"))
set.seed(23)

suppressPackageStartupMessages({
  library(tidyverse)
  library(MASS)
  library(rio)
  library(emmeans)
  library(ggplot2)
  library(patchwork)
})

fit_polr_safe <- function(formula, data) {
  suppressWarnings(
    MASS::polr(formula, data = data, Hess = TRUE, method = "logistic", model = TRUE)
  )
}

boot_expected_ci <- function(data, formula, grid, B = BOOT_REPS, seed = 23) {
  set.seed(seed)
  n <- nrow(data)
  boot_mat <- matrix(NA_real_, nrow = nrow(grid), ncol = B)

  for (b in seq_len(B)) {
    idx <- sample.int(n, size = n, replace = TRUE)
    dat_b <- droplevels(data[idx, , drop = FALSE])

    fit_b <- try(
      suppressWarnings(MASS::polr(formula, data = dat_b, Hess = FALSE, method = "logistic")),
      silent = TRUE
    )

    if (!inherits(fit_b, "try-error")) {
      vals <- try(expected_from_polr(fit_b, grid), silent = TRUE)
      if (!inherits(vals, "try-error")) {
        boot_mat[, b] <- vals
      }
    }
  }

  tibble(
    ymin = apply(boot_mat, 1, quantile, probs = 0.025, na.rm = TRUE),
    ymax = apply(boot_mat, 1, quantile, probs = 0.975, na.rm = TRUE)
  )
}

expected_from_polr <- function(model, newdata) {
  probs <- as.data.frame(predict(model, newdata = newdata, type = "probs"))
  probs[, "Never"] * 1 +
    probs[, "1-11 d/year"] * 2 +
    probs[, "1-3 d/month"] * 3 +
    probs[, "Weekly or greater"] * 4
}

coef_table_fig <- function(model) {
  ct <- coef(summary(model))
  tibble(
    term = rownames(ct),
    b = ct[, "Value"],
    SE = ct[, "Std. Error"],
    z = ct[, "t value"],
    p = 2 * pnorm(abs(ct[, "t value"]), lower.tail = FALSE)
  )
}

simple_slopes_fig <- function(model, var_name) {
  trends <- emmeans::emtrends(model, specs = ~ Sex, var = var_name, mode = "latent")
  bind_rows(
    as_tibble(as.data.frame(summary(trends, infer = c(TRUE, TRUE)))) %>%
      transmute(
        contrast = as.character(Sex),
        b = !!sym(paste0(var_name, ".trend")),
        p = p.value
      ),
    as_tibble(as.data.frame(summary(pairs(trends), infer = c(TRUE, TRUE)))) %>%
      transmute(
        contrast = as.character(contrast),
        b = estimate,
        p = p.value
      )
  )
}

fmt_b2 <- function(x) formatC(x, format = "f", digits = 2)

dfbetas_results_for_figures <- results_dfbetas
slopes_results_for_figures <- results_slopes

if (!"hemisphere" %in% names(slopes_results_for_figures) && nrow(slopes_results_for_figures) > 0) {
  rows_per_hemi <- nrow(slopes_results_for_figures) / 2
  if (rows_per_hemi == floor(rows_per_hemi)) {
    slopes_results_for_figures <- slopes_results_for_figures %>%
      mutate(
        hemisphere = rep(c("Left", "Right"), each = rows_per_hemi),
        .before = 1
      )
  } else {
    slopes_results_for_figures <- slopes_results_for_figures %>%
      mutate(hemisphere = "Right", .before = 1)
  }
}

if (!"analysis" %in% names(slopes_results_for_figures)) {
  slopes_results_for_figures <- slopes_results_for_figures %>%
    mutate(analysis = NA_character_)
}

if (!"contrast" %in% names(slopes_results_for_figures)) {
  slopes_results_for_figures <- slopes_results_for_figures %>%
    mutate(contrast = NA_character_)
}

plus_after_dfbetas <- function(model_name) {
  out <- dfbetas_results_for_figures %>%
    filter(.data$model == model_name, .data$cutoff_label == "max |DFBETAS| > .20")
  nrow(out) > 0 && isTRUE(out$p[1] < 0.05)
}

sig_marks <- function(p, plus = FALSE) {
  paste0(ifelse(isTRUE(p < 0.05), "*", ""), ifelse(plus, "+", ""))
}

main_effect_label <- function(model, term, model_name) {
  x <- coef_table_fig(model) %>% filter(.data$term == !!term)
  paste0("\U0001D44F = ", fmt_b2(x$b), sig_marks(x$p, FALSE))
}

sex_slope_label <- function(model, var_name, model_name) {
  slopes <- simple_slopes_fig(model, var_name)
  f <- slopes %>% filter(.data$contrast == "F")
  m <- slopes %>% filter(.data$contrast == "M")
  diff <- slopes %>% filter(.data$contrast == "F - M")
  hemisphere <- ifelse(grepl("^left", model_name), "Left", "Right")
  conservative_slope_sig <- function(contrast_label) {
    if (hemisphere != "Right") return(FALSE)
    out <- slopes_results_for_figures %>%
      filter(
        .data$hemisphere == hemisphere,
        .data$analysis == "After max |DFBETAS| > 2/sqrt(N) removal",
        .data$contrast == contrast_label
      )
    nrow(out) > 0 && isTRUE(out$p[1] < 0.05)
  }
  f_plus <- conservative_slope_sig("F")
  m_plus <- conservative_slope_sig("M")
  paste0(
    "F: \U0001D44F = ", fmt_b2(f$b), sig_marks(f$p, f_plus), "\n",
    "M: \U0001D44F = ", fmt_b2(m$b), sig_marks(m$p, m_plus), "\n",
    "F-M: \U0001D44F = ", fmt_b2(diff$b), ifelse(isTRUE(diff$p < 0.05), "*", "")
  )
}

HCP_dat <- rio::import(DATA_PATH)

analysis_dat <- HCP_dat %>%
  mutate(
    Subject = as.character(Subject),
    Sex = factor(trimws(Gender), levels = c("F", "M")),
    binge_freq_rc = case_when(
      SSAGA_Alc_12_Frq_5plus == 5 ~ 1L,
      SSAGA_Alc_12_Frq_5plus == 4 ~ 2L,
      SSAGA_Alc_12_Frq_5plus == 3 ~ 3L,
      SSAGA_Alc_12_Frq_5plus %in% c(1L, 2L) ~ 4L,
      TRUE ~ NA_integer_
    ),
    binge_freq_rc_f = factor(
      binge_freq_rc,
      levels = c(1, 2, 3, 4),
      labels = c("Never", "1-11 d/year", "1-3 d/month", "Weekly or greater"),
      ordered = TRUE
    )
  ) %>%
  transmute(
    row_id = row_number(),
    Subject,
    Sex,
    y = binge_freq_rc_f,
    left_hpc_vs = as.numeric(limbic_Left),
    right_hpc_vs = as.numeric(limbic_Right),
    ICV = as.numeric(FS_IntraCranial_Vol),
    Age = as.numeric(Age_in_Yrs)
  ) %>%
  filter(
    is.finite(left_hpc_vs),
    is.finite(right_hpc_vs),
    is.finite(ICV),
    is.finite(Age),
    !is.na(y),
    !is.na(Sex)
  ) %>%
  mutate(
    c_left_hpc_vs = as.numeric(scale(left_hpc_vs)),
    c_right_hpc_vs = as.numeric(scale(right_hpc_vs)),
    c_ICV = as.numeric(scale(ICV))
  ) %>%
  droplevels()

left_dat <- analysis_dat %>%
  dplyr::select(row_id, Subject, Sex, y, c_left_hpc_vs, c_ICV) %>%
  droplevels()

right_dat <- analysis_dat %>%
  dplyr::select(row_id, Subject, Sex, y, c_right_hpc_vs, c_ICV) %>%
  droplevels()

left_main <- fit_polr_safe(y ~ c_ICV + c_left_hpc_vs, data = left_dat)
right_main <- fit_polr_safe(y ~ c_ICV + c_right_hpc_vs, data = right_dat)
left_int <- fit_polr_safe(y ~ c_ICV + c_left_hpc_vs * Sex, data = left_dat)
right_int <- fit_polr_safe(y ~ c_ICV + c_right_hpc_vs * Sex, data = right_dat)

left_main_label <- main_effect_label(left_main, "c_left_hpc_vs", "left_main")
right_main_label <- main_effect_label(right_main, "c_right_hpc_vs", "right_main")
left_sex_label <- sex_slope_label(left_int, "c_left_hpc_vs", "left_interaction")
right_sex_label <- sex_slope_label(right_int, "c_right_hpc_vs", "right_interaction")

left_main_grid <- tibble(
  c_ICV = 0,
  c_left_hpc_vs = seq(min(left_dat$c_left_hpc_vs), max(left_dat$c_left_hpc_vs), length.out = 200)
)
left_main_grid$expected_y <- expected_from_polr(left_main, left_main_grid)
left_main_grid <- bind_cols(
  left_main_grid,
  boot_expected_ci(left_dat, y ~ c_ICV + c_left_hpc_vs, left_main_grid, seed = 3101)
)

right_main_grid <- tibble(
  c_ICV = 0,
  c_right_hpc_vs = seq(min(right_dat$c_right_hpc_vs), max(right_dat$c_right_hpc_vs), length.out = 200)
)
right_main_grid$expected_y <- expected_from_polr(right_main, right_main_grid)
right_main_grid <- bind_cols(
  right_main_grid,
  boot_expected_ci(right_dat, y ~ c_ICV + c_right_hpc_vs, right_main_grid, seed = 3102)
)

left_sex_grid <- expand_grid(
  c_ICV = 0,
  Sex = levels(left_dat$Sex),
  c_left_hpc_vs = seq(min(left_dat$c_left_hpc_vs), max(left_dat$c_left_hpc_vs), length.out = 200)
)
left_sex_grid$expected_y <- expected_from_polr(left_int, left_sex_grid)
left_sex_grid <- bind_cols(
  left_sex_grid,
  boot_expected_ci(left_dat, y ~ c_ICV + c_left_hpc_vs * Sex, left_sex_grid, seed = 3103)
)

right_sex_grid <- expand_grid(
  c_ICV = 0,
  Sex = levels(right_dat$Sex),
  c_right_hpc_vs = seq(min(right_dat$c_right_hpc_vs), max(right_dat$c_right_hpc_vs), length.out = 200)
)
right_sex_grid$expected_y <- expected_from_polr(right_int, right_sex_grid)
right_sex_grid <- bind_cols(
  right_sex_grid,
  boot_expected_ci(right_dat, y ~ c_ICV + c_right_hpc_vs * Sex, right_sex_grid, seed = 3104)
)

y_scale <- scale_y_continuous(
  breaks = 1:4,
  labels = c("Never", "1-11 d/year", "1-3 d/month", "Weekly or greater"),
  limits = c(0.75, 4.25)
)

base_assoc_theme <- theme_bw(base_size = 19) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 25, margin = margin(b = 8)),
    axis.title = element_text(size = 22),
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 12)),
    axis.text = element_text(size = 20, color = "gray20"),
    legend.title = element_text(size = 21),
    legend.text = element_text(size = 20),
    legend.key.height = unit(0.28, "in"),
    legend.key.width = unit(0.38, "in"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.7),
    plot.margin = margin(14, 22, 14, 22)
  )

fig_left_pooled <- ggplot(left_dat, aes(x = c_left_hpc_vs, y = as.numeric(y))) +
  geom_jitter(height = 0.08, width = 0, alpha = 0.38, size = 2.3, color = "black") +
  geom_ribbon(
    data = left_main_grid,
    aes(x = c_left_hpc_vs, y = expected_y, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    alpha = 0.18,
    fill = "#1f5fd2"
  ) +
  geom_line(data = left_main_grid, aes(y = expected_y), linewidth = 1.9, color = "#1f5fd2") +
  annotate("text", x = 4.15, y = 0.88, label = left_main_label, hjust = 1, vjust = 0,
           size = 5.8, fontface = "bold") +
  y_scale +
  coord_cartesian(xlim = c(-2.6, 4.3)) +
  labs(
    x = "HPC-VS connectivity\n(left, z-scored)",
    y = "Past-year 5+ drinking frequency",
    title = "A. Left hemisphere"
  ) +
  base_assoc_theme

fig_right_pooled <- ggplot(right_dat, aes(x = c_right_hpc_vs, y = as.numeric(y))) +
  geom_jitter(height = 0.08, width = 0, alpha = 0.38, size = 2.3, color = "black") +
  geom_ribbon(
    data = right_main_grid,
    aes(x = c_right_hpc_vs, y = expected_y, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    alpha = 0.18,
    fill = "#1f5fd2"
  ) +
  geom_line(data = right_main_grid, aes(y = expected_y), linewidth = 1.9, color = "#1f5fd2") +
  annotate("text", x = 4.15, y = 0.88, label = right_main_label, hjust = 1, vjust = 0,
           size = 5.8, fontface = "bold") +
  y_scale +
  coord_cartesian(xlim = c(-2.6, 4.3)) +
  labs(
    x = "HPC-VS connectivity\n(right, z-scored)",
    y = "Past-year 5+ drinking frequency",
    title = "B. Right hemisphere"
  ) +
  base_assoc_theme +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

fig_assoc_pooled <- fig_left_pooled + fig_right_pooled + plot_layout(widths = c(1, 1))

fig_left_by_sex <- ggplot(left_dat, aes(x = c_left_hpc_vs, y = as.numeric(y), color = Sex)) +
  geom_jitter(height = 0.08, width = 0, alpha = 0.46, size = 2.3) +
  geom_ribbon(
    data = left_sex_grid,
    aes(x = c_left_hpc_vs, y = expected_y, ymin = ymin, ymax = ymax, fill = Sex),
    inherit.aes = FALSE,
    alpha = 0.14,
    color = NA
  ) +
  geom_line(data = left_sex_grid, aes(y = expected_y), linewidth = 1.9) +
  annotate("text", x = 4.15, y = 0.88, label = left_sex_label, hjust = 1, vjust = 0,
           size = 5.1, fontface = "bold", lineheight = 0.95, color = "black") +
  scale_color_manual(values = c("F" = "#1f5fd2", "M" = "#f39c12")) +
  scale_fill_manual(values = c("F" = "#1f5fd2", "M" = "#f39c12"), guide = "none") +
  y_scale +
  coord_cartesian(xlim = c(-2.6, 4.3)) +
  labs(
    x = "HPC-VS connectivity\n(left, z-scored)",
    y = "Past-year 5+ drinking frequency",
    title = "A. Left hemisphere",
    color = "Sex"
  ) +
  base_assoc_theme

fig_right_by_sex <- ggplot(right_dat, aes(x = c_right_hpc_vs, y = as.numeric(y), color = Sex)) +
  geom_jitter(height = 0.08, width = 0, alpha = 0.46, size = 2.3) +
  geom_ribbon(
    data = right_sex_grid,
    aes(x = c_right_hpc_vs, y = expected_y, ymin = ymin, ymax = ymax, fill = Sex),
    inherit.aes = FALSE,
    alpha = 0.14,
    color = NA
  ) +
  geom_line(data = right_sex_grid, aes(y = expected_y), linewidth = 1.9) +
  annotate("text", x = 4.15, y = 0.88, label = right_sex_label, hjust = 1, vjust = 0,
           size = 5.1, fontface = "bold", lineheight = 0.95, color = "black") +
  scale_color_manual(values = c("F" = "#1f5fd2", "M" = "#f39c12")) +
  scale_fill_manual(values = c("F" = "#1f5fd2", "M" = "#f39c12"), guide = "none") +
  y_scale +
  coord_cartesian(xlim = c(-2.6, 4.3)) +
  labs(
    x = "HPC-VS connectivity\n(right, z-scored)",
    y = "Past-year 5+ drinking frequency",
    title = "B. Right hemisphere",
    color = "Sex"
  ) +
  base_assoc_theme +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

fig_assoc_by_sex <- fig_left_by_sex + fig_right_by_sex + plot_layout(widths = c(1, 1), guides = "collect") &
  theme(legend.position = "right")

print(fig_assoc_pooled)
knitr::asis_output("\\clearpage")
print(fig_assoc_by_sex)
```

# Final Figure 5

This section creates the diagnosis-group HPC-VS connectivity figure reported in
the manuscript.

```{r final-figure-5, fig.width=12.5, fig.height=7.0, fig.cap="HPC-VS connectivity by alcohol diagnosis group."}
set.seed(23)

suppressPackageStartupMessages({
  library(tidyverse)
  library(rio)
  library(ggplot2)
  library(patchwork)
})

fmt_p_fig <- function(p) {
  ifelse(p < 0.001, "< .001", paste0("= ", sub("^0", "", formatC(p, format = "f", digits = 3))))
}

HCP_dat <- rio::import(DATA_PATH)

diagnosis_plot_dat <- HCP_dat %>%
  mutate(
    Subject = as.character(Subject),
    Sex = factor(trimws(Gender), levels = c("F", "M")),
    alcohol_abuse_dx = case_when(
      SSAGA_Alc_D4_Ab_Dx == 5 ~ TRUE,
      SSAGA_Alc_D4_Ab_Dx == 1 ~ FALSE,
      TRUE ~ NA
    ),
    alcohol_dependence_dx = case_when(
      SSAGA_Alc_D4_Dp_Dx == 5 ~ TRUE,
      SSAGA_Alc_D4_Dp_Dx == 1 ~ FALSE,
      TRUE ~ NA
    ),
    DxGroup = factor(
      case_when(
        alcohol_dependence_dx ~ "Dependence",
        alcohol_abuse_dx ~ "Abuse",
        alcohol_abuse_dx == FALSE & alcohol_dependence_dx == FALSE ~ "No AUD",
        TRUE ~ NA_character_
      ),
      levels = c("No AUD", "Abuse", "Dependence")
    )
  ) %>%
  transmute(
    Subject,
    Sex,
    DxGroup,
    left_hpc_vs = as.numeric(limbic_Left),
    right_hpc_vs = as.numeric(limbic_Right),
    ICV = as.numeric(FS_IntraCranial_Vol)
  ) %>%
  filter(
    is.finite(left_hpc_vs),
    is.finite(right_hpc_vs),
    is.finite(ICV),
    !is.na(Sex),
    !is.na(DxGroup)
  ) %>%
  mutate(
    c_left_hpc_vs = as.numeric(scale(left_hpc_vs)),
    c_right_hpc_vs = as.numeric(scale(right_hpc_vs)),
    c_ICV = as.numeric(scale(ICV))
  ) %>%
  droplevels()

left_conn_reduced <- lm(c_left_hpc_vs ~ Sex + c_ICV, data = diagnosis_plot_dat)
left_conn_full <- lm(c_left_hpc_vs ~ DxGroup + Sex + c_ICV, data = diagnosis_plot_dat)
right_conn_reduced <- lm(c_right_hpc_vs ~ Sex + c_ICV, data = diagnosis_plot_dat)
right_conn_full <- lm(c_right_hpc_vs ~ DxGroup + Sex + c_ICV, data = diagnosis_plot_dat)

left_p <- anova(left_conn_reduced, left_conn_full)$`Pr(>F)`[2]
right_p <- anova(right_conn_reduced, right_conn_full)$`Pr(>F)`[2]

diagnosis_long <- diagnosis_plot_dat %>%
  dplyr::select(Subject, DxGroup, c_left_hpc_vs, c_right_hpc_vs) %>%
  pivot_longer(
    cols = c(c_left_hpc_vs, c_right_hpc_vs),
    names_to = "hemisphere",
    values_to = "connectivity"
  ) %>%
  mutate(
    hemisphere = factor(
      hemisphere,
      levels = c("c_left_hpc_vs", "c_right_hpc_vs"),
      labels = c(
        paste0("A. Left hemisphere\nANCOVA p ", fmt_p_fig(left_p)),
        paste0("B. Right hemisphere\nANCOVA p ", fmt_p_fig(right_p))
      )
    )
  )

fig_diagnosis <- ggplot(diagnosis_long, aes(x = DxGroup, y = connectivity, fill = DxGroup)) +
  geom_violin(width = 0.92, alpha = 0.36, color = NA, trim = FALSE) +
  geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.75, color = "gray20", linewidth = 0.8) +
  geom_jitter(width = 0.08, height = 0, alpha = 0.42, size = 2.1, color = "gray15") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4.6, fill = "white", color = "black") +
  facet_wrap(~ hemisphere, nrow = 1) +
  scale_fill_manual(values = c("No AUD" = "#89e6a8", "Abuse" = "#87bff5", "Dependence" = "#f6b737"), guide = "none") +
  labs(
    title = "HPC-VS Connectivity by Alcohol Diagnosis Group",
    x = "Lifetime DSM-IV alcohol diagnosis group",
    y = "HPC-VS connectivity (z-scored)"
  ) +
  theme_bw(base_size = 24) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 34, margin = margin(b = 10)),
    strip.background = element_rect(fill = "gray85", color = "gray20", linewidth = 0.9),
    strip.text = element_text(face = "bold", size = 25, lineheight = 0.95),
    axis.title = element_text(size = 25),
    axis.text = element_text(size = 22, color = "gray20"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.7),
    plot.margin = margin(14, 22, 14, 22)
  )

fig_diagnosis
```
